---
title: "Random stuff re: stability properties for MAR(1) models"
output: pdf_document
fontsize: 10pt
geometry: margin=1.25in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\begin{center}
\textbf{Mark D. Scheuerell} \\
\text{Northwest Fisheries Science Center} \\
\text{National Oceanic and Atmospheric Administration} \\
\text{Seattle, WA USA} \\
\text{mark.scheuerell@noaa.gov}
\end{center}

# Example via decomposition

Now let's fit the MARSS model to the L Washington plankton data after we have removed any seasonal effect via decomposition ($i.e.$, removed the monthly means), rather than fitting some dummy covariates for season.

```{r decomp_data}
yy <- dat[,all_names]
yy[yy==0] <- NA
## log-transform variates
## data must also be n x T, so transpose
yy <- t(log(yy))
## subtract monthly means
for(i in 1:12) {
  idx <- seq(from=i, by=12, length.out=TT/12)
  mm <- apply(yy[,idx], 1, mean, na.rm=TRUE)
  yy[,idx] <- yy[,idx] - mm
}
## plot the decomposed data
matplot(t(yy), type="l")
```

Now we can fit the model. Note that I set $\mathbf{C}$ and $\mathbf{c}$ to `"zero"` because we have eliminated the seasonal effect.

```{r fit_mod_2, cache=TRUE}
mod_list$C <- "zero"
mod_list$c <- "zero"
## fit model
mod <- MARSS(yy, model=mod_list, control=list(maxit=5000, safe=TRUE))
## get B matrix
BB <- coef(mod, type="matrix")$B
round(BB,2)
## calc pi_B
det(BB)^2
## Ives adjustment for other community comparisons
det(BB)^(2/nn)
```
